---
title: "Cook County Electricity Demand Time Series Modeling - Simple Methods"
output: html_document
---

# Load libraries

```{r,warning=F,message=F}
library(forecast)
library(tseries)
library(ggplot2)
library(dplyr)

Sys.setlocale("LC_ALL", "en_US.UTF-8")
```

# Load Data

```{r}
df_1819 <- readRDS("data/train_test_split/df_1819.rds")
df_2020   <- readRDS("data/train_test_split/df_20.rds")
```

# Modeling

### Simple Methods (Baselines)

```{r}
# train
ts_1819 <- ts(df_1819$electricity_demand, frequency = 24)

# test
ts_2020 <- ts(
  df_2020$electricity_demand,
  frequency = 24,
  start = time(ts_1819)[length(ts_1819)] + 1/frequency(ts_1819)
)

# seasonal naive
fit_snaive <- snaive(ts_1819, h = length(ts_2020))

# plot
autoplot(fit_snaive) +
  autolayer(ts_2020, series = "Actual Demand (Jan 2020)", color = "red") +
  coord_cartesian(xlim = c(max(time(ts_1819)) - 10, max(time(ts_2020)))) +
  labs(title = "Seasonal Naive Forecast vs Actual",
       x = "Time (Days)",
       y = "Electricity Demand") +
  theme_minimal()

# accuracy
acc_snaive <- accuracy(fit_snaive, ts_2020)
print(acc_snaive)

checkresiduals(fit_snaive)
```

The plot compares this repeated daily profile against the January 2020 actual demand, so any persistent gap indicates level changes or calendar/weather-driven shifts not present at the end of 2019. Quantitatively, the test MAPE is about 5.25%, which is a reasonable benchmark, but the very high residual autocorrelation suggests substantial remaining structure beyond simple daily seasonality.

```{r}
# Fit Mean Forecast model
fit_mean <- meanf(ts_1819, h = length(ts_2020))

# Plot the Mean forecast
autoplot(fit_mean, PI = FALSE, series = "Mean Forecast") +
  autolayer(ts_2020, series = "Actual Demand (Jan 2020)", color = "red") +
  coord_cartesian(xlim = c(max(time(ts_1819)) - 10, max(time(ts_2020)))) +
  labs(title = "Mean Forecast vs Actual",
       x = "Time (Days)",
       y = "Electricity Demand") +
  theme_minimal() +
  scale_color_manual(values = c("Actual Demand (Jan 2020)" = "red", "Mean Forecast" = "blue"))

# Evaluate accuracy bypassing ts index mismatch
acc_mean <- accuracy(fit_mean, df_2020$electricity_demand)
print(acc_mean)

```

```{r}
# Fit STLF model
fit_stlf <- stlf(ts_1819, method = "naive", h = length(ts_2020))

# Plot the STLF forecast
autoplot(fit_stlf, PI = FALSE, series = "STLF Forecast") +
  autolayer(ts_2020, series = "Actual Demand (Jan 2020)", color = "red") +
  coord_cartesian(xlim = c(max(time(ts_1819)) - 10, max(time(ts_2020)))) +
  labs(title = "STLF Forecast vs Actual",
       x = "Time (Days)",
       y = "Electricity Demand") +
  theme_minimal() +
  scale_color_manual(values = c("Actual Demand (Jan 2020)" = "red", "STLF Forecast" = "blue"))

# Evaluate accuracy
acc_stlf <- accuracy(fit_stlf, df_2020$electricity_demand)
print(acc_stlf)

```

```{r}
#Time Series Linear Model (TSLM) & Diagnostics
library(lmtest)

# Fit TSLM with trend and daily seasonality
fit_tslm <- tslm(ts_1819 ~ trend + season)
summary(fit_tslm)

autoplot(window(ts_1819, start = time(ts_1819)[length(ts_1819) - 336])) +
  autolayer(window(fitted(fit_tslm), start = time(ts_1819)[length(ts_1819) - 336]), 
            series = "Fitted (TSLM)", size = 0.8) +
  labs(title = "TSLM: Actual vs. Fitted (Last 14 Days of Training)",
       x = "Time",
       y = "Electricity Demand") +
  theme_minimal() +
  scale_color_manual(values = c("Data" = "black", "Fitted (TSLM)" = "red"))

# Residual Diagnostics
res_tslm <- residuals(fit_tslm)

# Plot Residuals Time Plot
autoplot(res_tslm) +
  labs(title = "Residuals from TSLM", y = "Residuals") +
  theme_minimal()

# Plot ACF of Residuals
ggAcf(res_tslm, lag.max = 48) +
  labs(title = "ACF of TSLM Residuals")

# Plot Histogram of Residuals
hist(res_tslm, breaks = 50, col = "lightblue", 
     main = "Histogram of TSLM Residuals", xlab = "Residuals")

# Ljung-Box Test
Box.test(res_tslm, lag = 24, type = "Ljung-Box")

# Breusch-Godfrey Test
bgtest(fit_tslm)

# Forecast and Plot
fcast_tslm <- forecast(fit_tslm, h = length(df_2020$electricity_demand))

autoplot(fcast_tslm, PI = FALSE, series = "TSLM Forecast") +
  autolayer(ts_2020, series = "Actual Demand (Jan 2020)") +
  coord_cartesian(xlim = c(max(time(ts_1819)) - 10, max(time(ts_2020)))) +
  labs(title = "TSLM Forecast vs Actual",
       x = "Time (Days)",
       y = "Electricity Demand") +
  theme_minimal() +
  scale_color_manual(values = c("Actual Demand (Jan 2020)" = "red", "TSLM Forecast" = "blue"))

acc_tslm <- accuracy(fcast_tslm, df_2020$electricity_demand)
print(acc_tslm)
```

For visualization purposes, only the last 14 days of the testing period are shown. This window includes multiple daily and weekly seasonal cycles, making it sufficient to evaluate forecast alignment while maintaining visual clarity.
