---
title: "Cook County Electricity Demand Time Series Modeling Handbook"
output: html_document
---

```{r,warning=F,message=F}
library(forecast)
library(tseries)
library(ggplot2)
library(dplyr)
library(lubridate)
library(future.apply)
Sys.setlocale("LC_ALL", "en_US.UTF-8")
```

```{r}
# load data
df_1819 <- readRDS("data/train_test_split/df_1819.rds")
df_20   <- readRDS("data/train_test_split/df_20.rds")

# convert into expected datatype
ts_1819 <- ts(df_1819$electricity_demand, frequency = 24)
ts_20   <- ts(df_20$electricity_demand, frequency = 24)
```

## Train - Validation - Test Framework

1.  Data is split into train & test data.
2.  Models will be trained and tuned using `train data`. Different or tailored train - validation structure might be used accordingly. This file provide a sample structure.
3.  All the models will be evaluated on the forecasting performance on the test data.
4.  Metrics consider AICc, RMSE, MAE and MAPE.

**Note**: Train-validation procedure can be different, while the test output should strictly follow the same pattern.

### Data Size & hyperparameters for train - validation

| parameter | meaning                 |
|-----------|-------------------------|
| k         | initial training window |
| H         | forecast horizon        |
| stride    | sliding steps           |
| p         | seasonal period         |
| I         | number of iterations    |

```{r}
k      <- 24 * 365       # initial training window: 1 year
# k      <- 24 * 365 * 1.5 # 1.5 year
# H      <- 24             # validation horizon: 24 hours (day-ahead)
H      <- 24 * 7         # week-ahead
stride <- 24*7

p <- frequency(ts_1819)
n <- length(ts_1819)
I <- floor((n - k - H)/stride)  # iterations count based on stride
I                               # check number of iterations for the computing time estimate
```

### Cross Validation Framework

ETS auto fit as example

#### Setting

```{r}
# Containers
make_store <- function(I, H) list(err = matrix(NA_real_, I, H), aicc = rep(NA_real_, I))
ets_exp <- make_store(I, H)
ets_sld <- make_store(I, H)

# Model fit function
fit_ets <- function(y_train) {
  ets(y_train)   # auto ETS
}

# Helper: index -> ts time (works for any ts frequency)
idx2time <- function(x_ts, idx) {
  tp <- tsp(x_ts)       # c(start, end, frequency)
  st <- tp[1]
  p  <- tp[3]
  st + (idx - 1) / p
}

# Metrics helpers (per-iteration)
mae_iter  <- function(err_mat) apply(abs(err_mat), 1, mean, na.rm = TRUE)
rmse_iter <- function(err_mat) apply(err_mat^2, 1, function(x) sqrt(mean(x, na.rm = TRUE)))
```

#### CV loop

```{r}
# Use multisession for Windows; multicore for Mac/Linux is faster.
plan(multisession, workers = 16 )

# One-iteration worker
one_iter <- function(i, x_ts, k, H, stride, n) {
  
  # train - valid window split
  exp_end   <- k + (i - 1) * stride
  sld_start <- 1 + (i - 1) * stride
  sld_end   <- exp_end

  val_start <- exp_end + 1
  val_end   <- min(val_start + H - 1, n)
  L <- val_end - val_start + 1

  exp_end_t   <- idx2time(x_ts, exp_end)
  sld_start_t <- idx2time(x_ts, sld_start)
  sld_end_t   <- exp_end_t
  val_start_t <- idx2time(x_ts, val_start)
  val_end_t   <- idx2time(x_ts, val_end)

  y_exp   <- window(x_ts, end   = exp_end_t)
  y_sld   <- window(x_ts, start = sld_start_t, end = sld_end_t)
  y_valid <- window(x_ts, start = val_start_t, end = val_end_t)

  # expanding
  m_exp <- fit_ets(y_exp)
  f_exp <- forecast(m_exp, h = H)
  err_exp <- rep(NA_real_, H)
  err_exp[1:L] <- as.numeric(f_exp$mean[1:L] - y_valid)

  # sliding
  m_sld <- fit_ets(y_sld)
  f_sld <- forecast(m_sld, h = H)
  err_sld <- rep(NA_real_, H)
  err_sld[1:L] <- as.numeric(f_sld$mean[1:L] - y_valid)
  
  # returns a list with exp/sld errors + aicc
  list(
    err_exp  = err_exp,
    aicc_exp = m_exp$aicc,
    err_sld  = err_sld,
    aicc_sld = m_sld$aicc
  )
}
```

```{r}
# Run CV in parallel
# future.seed=TRUE ensures reproducibility if any randomness appears
res <- future_lapply(
  X = seq_len(I),
  FUN = one_iter,
  x_ts = ts_1819, k = k, H = H, stride = stride, n = n,
  future.seed = TRUE
)
```

#### Evaluate on validation data

```{r}
# Collect results into same store structure
ets_exp <- make_store(I, H)
ets_sld <- make_store(I, H)

for (i in seq_len(I)) {
  ets_exp$err[i, ]  <- res[[i]]$err_exp
  ets_exp$aicc[i]   <- res[[i]]$aicc_exp
  ets_sld$err[i, ]  <- res[[i]]$err_sld
  ets_sld$aicc[i]   <- res[[i]]$aicc_sld
}
```

```{r}
# Summaries (same as your template)
mae_ets_exp_mean  <- mean(mae_iter(ets_exp$err),  na.rm = TRUE)
rmse_ets_exp_mean <- mean(rmse_iter(ets_exp$err), na.rm = TRUE)
aicc_ets_exp_mean <- mean(ets_exp$aicc,           na.rm = TRUE)

mae_ets_sld_mean  <- mean(mae_iter(ets_sld$err),  na.rm = TRUE)
rmse_ets_sld_mean <- mean(rmse_iter(ets_sld$err), na.rm = TRUE)
aicc_ets_sld_mean <- mean(ets_sld$aicc,           na.rm = TRUE)

df_summary <- data.frame(
  Model     = c("ETS Expanding", "ETS Sliding"),
  Mean_MAE  = c(mae_ets_exp_mean,  mae_ets_sld_mean),
  Mean_RMSE = c(rmse_ets_exp_mean, rmse_ets_sld_mean),
  Mean_AICc = c(aicc_ets_exp_mean, aicc_ets_sld_mean)
)

df_summary
```

### Forecast output on Test Data

```{r}
modeling_method <- "ETS_Sample"
forecast_path  <- paste0("data/test_forecast/", modeling_method, "_20Jan.csv")

h_fc   <- nrow(df_20)
ts_out <- df_20$timestamp

fit_ets <- ets(ts_1819)
fc_ets  <- forecast(fit_ets, h = h_fc)
y_pred  <- as.numeric(fc_ets$mean)

test_forecast <- tibble(
  timestamp = ts_out,
  y_pred    = y_pred
)
write.csv(test_forecast, forecast_path, row.names = FALSE)
```
