---
title: "Cook County Electricity Demand Time Series Modeling - SARIMAX"
output: html_document
---

```{r,warning=F,message=F}
library(forecast)
library(readr)
library(tseries)
library(ggplot2)
library(dplyr)
library(lubridate)
library(future.apply)
Sys.setlocale("LC_ALL", "en_US.UTF-8")
```

```{r}
# load data
df_1819 <- readRDS("data/train_test_split/df_1819.rds")
df_1819 <- df_1819[-1, ]
df_20   <- readRDS("data/train_test_split/df_20.rds")

cutoff <- ymd_hms("2020-01-01 00:00:00", tz="UTC")
```



### Covariate: Weather + Month dummy

```{r}
weather_path <- "data/raw/ORD_weather.csv"  

df_weather <- read.csv(weather_path, stringsAsFactors = FALSE) %>%
  filter(valid != "valid") %>%
  mutate(
    timestamp = ymd_hm(valid, tz = "UTC"),
    tmpf = parse_number(tmpf),
    dwpf = parse_number(dwpf),
    sped = parse_number(sped),
    p01m = parse_number(p01m),
    ts_hour_utc = round_date(timestamp, "hour")
  )
head(df_weather,2)
```
- station: air monitoring station.
- valid: utc timestamp as string
- tmpf: air temperature [F]
- dwpf: dew point [F], the temperature to which air must be cooled to become saturated with water vapor (
 relative humidity)
- sped: wind speed [mph]
- p01m: 1 hour precipitation [mm]


**tidy up the weaather df to be aligned with elec demand df**
```{r}
# round to hour
df_weather_hr <- df_weather %>%
  group_by(ts_hour_utc) %>%
  summarise(
    across(c(tmpf, dwpf, sped, p01m),
           ~ mean(.x, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  arrange(ts_hour_utc)

# A complete hourly UTC time grid
full_hours <- tibble(
  ts_hour_utc = seq(min(df_weather_hr$ts_hour_utc),
                    max(df_weather_hr$ts_hour_utc),
                    by = "hour")
)

df_weather_full <- full_hours %>%
  left_join(df_weather_hr, by = "ts_hour_utc")

# Missing weather are linearly interpolated.
df_weather_full <- df_weather_full %>%
  arrange(ts_hour_utc) %>%
  mutate(t = as.numeric(ts_hour_utc)) %>%
  mutate(
    across(
      c(tmpf, dwpf, sped, p01m),
      ~ approx(t[!is.na(.x)],
               .x[!is.na(.x)],
               t,
               rule = 2)$y
    )
  ) %>%
  select(-t)
```

**Feature engineering on weather data**
```{r}
tmp_threshold <-50

df_weather_full <- df_weather_full %>%
  mutate(
    # time
    month = month(ts_hour_utc),
    
    # Heating Degree Hour & Cooling Degree Hour
    tmpf2 = (tmpf - tmp_threshold)^2,
    HDH   = pmax(tmp_threshold - tmpf, 0),
    HDH2  = HDH^2,
    CDH   = pmax(tmpf - tmp_threshold, 0),
    CDH2  = CDH^2,
    
    # season indicator
    is_winter = ifelse(month %in% c(12,1,2,3), 1, 0),
    is_summer = ifelse(month %in% c(6,7,8,9), 1, 0),
    
    # interaction
    winter_HDH  = HDH * is_winter,
    winter_wind = sped * is_winter,
    summer_CDH  = CDH * is_summer,
    summer_dew  = dwpf * is_summer
  )
```

**month dummy**
```{r}
df_weather_full$month_f <- factor(df_weather_full$month,
                                   levels = 1:12,
                                   labels = month.abb)
month_dummy <- model.matrix(~ month_f - 1, data = df_weather_full)

# drop one to avoid multilinearity
month_dummy <- month_dummy[, colnames(month_dummy) != "month_fJan", drop = FALSE]
```

**Final Xreg**

```{r}
df_xreg <- cbind(
  df_weather_full %>%
    select(ts_hour_utc, tmpf2, HDH, CDH, HDH2, CDH2, p01m, is_winter, is_summer, winter_HDH, winter_wind, summer_CDH, summer_dew) %>%
    rename(timestamp = ts_hour_utc),
  month_dummy
) %>%
  as.data.frame()

df_xreg_1819 <- df_xreg %>% filter(timestamp < cutoff)
df_xreg_20   <- df_xreg %>% filter(timestamp >= cutoff)
```


## Train - Validation - Test

### split train validation on 1819 data


```{r}
H_val <- 24 * 7 * 2  # 2 weeks

# 1) assure time sorting
df_1819      <- df_1819 %>% arrange(timestamp)
df_xreg_1819 <- df_xreg_1819 %>% arrange(timestamp)

# 2) Alignment check
stopifnot(nrow(df_1819) == nrow(df_xreg_1819))
stopifnot(all(df_1819$timestamp == df_xreg_1819$timestamp))

# 3) train/val index
n_1819 <- nrow(df_1819)
idx_train <- 1:(n_1819 - H_val)
idx_val   <- (n_1819 - H_val + 1):n_1819

y_train <- df_1819[idx_train, ]
y_val   <- df_1819[idx_val, ]

df_xreg_train <- df_xreg_1819[idx_train, ]
df_xreg_val   <- df_xreg_1819[idx_val, ]
```


**SARIMAX input dtype：ts + xreg matrix**
```{r}
y_col <- "electricity_demand"  

# ts: hourly, frequency=24 
y_train_ts <- ts(y_train[[y_col]], frequency = 24)
y_val_ts   <- ts(y_val[[y_col]],   frequency = 24)

# xreg: drop timestamp, as.matrix
xreg_train <- df_xreg_train %>% select(-timestamp) %>% as.matrix()
xreg_val   <- df_xreg_val   %>% select(-timestamp) %>% as.matrix()

stopifnot(length(y_train_ts) == nrow(xreg_train))
stopifnot(length(y_val_ts)   == nrow(xreg_val))
stopifnot(ncol(xreg_train)   == ncol(xreg_val))
stopifnot(all(colnames(xreg_train) == colnames(xreg_val)))
```


**plot to show xreg**
```{r fig.width=12, fig.height=6}
library(tidyr)

y_col <- "electricity_demand"

plot_base <- df_1819 %>%
  select(timestamp, all_of(y_col)) %>%
  inner_join(
    df_xreg_1819 %>%
      select(timestamp, tmpf2, winter_HDH, winter_wind, summer_CDH, summer_dew),
    by = "timestamp"
  ) %>%
  arrange(timestamp)

plot_df <- plot_base %>%
  pivot_longer(
    cols = -timestamp,
    names_to = "series",
    values_to = "value"
  ) %>%
  mutate(
    series = factor(
      series,
      levels = c(
        y_col,
        "tmpf^2,",
        "winter_HDH",
        "winter_wind",
        "summer_CDH",
        "summer_dew"
      )
    )
  )

ggplot(plot_df, aes(x = timestamp, y = value)) +
  geom_line() +
  facet_wrap(~series, ncol = 1, scales = "free_y") +
  theme_minimal() +
  labs(
    title = "Electricity Demand and Weather Features",
    x = NULL,
    y = NULL
  )
```
**multicollinearity**

```{r}
weather_features <- df_xreg_train %>%
  select(winter_HDH, winter_wind, summer_CDH, summer_dew)

library(usdm)
vif(weather_features)
```

summer_CDH and summer_dew can not be used at the same time.



### Assumptions made for the modeling

(S)ARIMAX models require future values of exogenous covariates when generating forecasts. In this application, weather features such as temperature and precipitation are used as an external regressor. Since reliable weather forecasts are available and generally highly accurate over short horizons, we assume that future temperature forecasts are equal to the realized temperature values. Therefore, weather features are treated as known when producing demand forecasts.

We assume the covariate can solve the heteroscedasticity issue.

The weather data is actually the records from ord station in Chicago, here we assume the weather condition in Chicago and Cook County are identical.

### Train - Validation


#### Create func for validation

```{r}
plot_residual_diag <- function(fit, timestamp) {
  res <- residuals(fit)
  par(mfrow = c(2,2), mar = c(4,4,2,1))
  plot(timestamp, res, type="l"); abline(h=0,lty=2)
  hist(res, breaks=50, prob=TRUE); lines(density(res, na.rm=TRUE))
  acf(res, lag.max=24, na.action=na.pass, ylim=c(-0.2,0.2))
  pacf(res, lag.max=24, na.action=na.pass)
  par(mfrow=c(1,1))
}

plot_residual_acf_longlag <- function(fit) {
  res <- residuals(fit)
  par(mfrow=c(2,1), mar=c(4,4,2,1))
  acf(res, lag.max=24*7*4, na.action=na.pass, ylim=c(-0.2,0.2))
  abline(v=c(7,14,21,28), col="red", lty=3)
  pacf(res, lag.max=24*7*4, na.action=na.pass)
  abline(v=c(7,14,21,28), col="red", lty=3)
  par(mfrow=c(1,1))
}
```


```{r}
calc_metrics <- function(actual, forecast_mean) {
  actual <- as.numeric(actual)
  forecast_mean <- as.numeric(forecast_mean)
  
  c(
    RMSE = sqrt(mean((actual - forecast_mean)^2)),
    MAE  = mean(abs(actual - forecast_mean)),
    MAPE = mean(abs((actual - forecast_mean) / actual)) * 100
  )
}

plot_train_valid_forecast <- function(
  y_train,           # data.frame, must contain timestamp
  y_train_ts,        # ts or numeric (train actual values)
  y_val,             # data.frame, must contain timestamp
  y_val_ts,          # ts or numeric (validation actual values)
  fc_obj,            # forecast object from forecast()
  ci_level = 2,      # which column of lower/upper to use (e.g., 2 for 95%)
  lookback_days = 7*12  # how many days before split to show
) {
  # --- forecast & actual vector ---
  yhat_valid <- as.numeric(fc_obj$mean)
  y_valid_vec <- as.numeric(y_val_ts)

  # --- align length (safety) ---
  h <- length(yhat_valid)

  # timestamps for validation/forecast must be length h
  ts_val <- y_val$timestamp[seq_len(h)]

  # --- 1. train actual ---
  train_df <- data.frame(
    timestamp = y_train$timestamp,
    value     = as.numeric(y_train_ts),
    type      = "Train"
  )

  # --- 2. validation actual (aligned) ---
  valid_df <- data.frame(
    timestamp = ts_val,
    value     = y_valid_vec[seq_len(h)],
    type      = "Validation"
  )

  # --- 3. forecast df (aligned) ---
  forecast_df <- data.frame(
    timestamp = ts_val,
    Forecast  = yhat_valid,
    Lower     = fc_obj$lower[seq_len(h), ci_level],
    Upper     = fc_obj$upper[seq_len(h), ci_level]
  )

  split_time <- min(ts_val)

  # --- 4. plot ---
  ggplot2::ggplot() +
    ggplot2::geom_line(
      data = train_df,
      ggplot2::aes(x = timestamp, y = value),
      color = "black", linewidth = 0.6
    ) +
    ggplot2::geom_line(
      data = valid_df,
      ggplot2::aes(x = timestamp, y = value),
      color = "black", linewidth = 0.6
    ) +
    ggplot2::geom_ribbon(
      data = forecast_df,
      ggplot2::aes(x = timestamp, ymin = Lower, ymax = Upper),
      fill = "pink", alpha = 0.3
    ) +
    ggplot2::geom_line(
      data = forecast_df,
      ggplot2::aes(x = timestamp, y = Forecast),
      color = "red", linewidth = 0.7
    ) +
    ggplot2::geom_vline(
      xintercept = split_time,
      linetype = "dashed", color = "blue"
    ) +
    ggplot2::theme_minimal() +
    ggplot2::labs(
      title = "Train + Validation Forecast",
      x = "Time",
      y = "Electricity Demand"
    ) +
    ggplot2::coord_cartesian(
      xlim = c(
        split_time - lubridate::days(lookback_days),
        max(ts_val)
      )
    )
}
```

#### Try 1

```{r}
weather_vars_subset1 <- c("tmpf2")
Xreg_input <- xreg_train[, weather_vars_subset1, drop = FALSE]

fit_sarimax_D1_1 <- Arima(
  y_train_ts,
  xreg = Xreg_input,
  order = c(4,0,3),
  seasonal = list(order = c(2,1,1), period = 24),
  method = "ML",
)

summary(fit_sarimax_D1_1)
```



**Validation**
```{r}
weather_vars <- weather_vars_subset1

xreg_val_subset <- xreg_val[, weather_vars, drop = FALSE]
fc_sarima_D1_1 <- forecast(fit_sarimax_D1_1, h = H_val, xreg = xreg_val_subset)
```


```{r}
calc_metrics(y_val_ts, fc_sarima_D1_1$mean)
```

```{r fig.width=12, fig.height=4}
plot_train_valid_forecast(
  y_train = y_train,
  y_train_ts = y_train_ts,
  y_val = y_val,
  y_val_ts = y_val_ts,
  fc_obj = fc_sarima_D1_1,
  ci_level = 2,
  lookback_days = 7*12
)
```


#### Try 2

```{r}
library(dplyr)

y_col <- "electricity_demand"
weather_vars_subset2 <- c("winter_HDH", "winter_wind", "CDH2", "p01m")

lm_df <- y_train %>%
  select(timestamp, y = all_of(y_col)) %>%
  left_join(
    df_xreg_train %>% select(timestamp, all_of(weather_vars_subset2)),
    by = "timestamp"
  )

fit_lm_log <- lm(
  log(y) ~ winter_HDH + winter_wind + CDH2 + p01m,
  data = lm_df
)

summary(fit_lm_log)
```



```{r}
weather_vars_subset2 <- c("winter_HDH","CDH2", "p01m")
Xreg_input <- xreg_train[, weather_vars_subset2, drop = FALSE]

fit_sarimax_D1_2 <- auto.arima(
  y_train_ts,
  xreg = Xreg_input,
  seasonal = TRUE,
  stepwise = TRUE,
  approximation = TRUE,
  d = 0, D = 1,
  max.p = 4, max.q = 4,
  max.P = 4, max.Q = 4
)

summary(fit_sarimax_D1_2)
```

**Check Residuals**
```{r fig.width=12, fig.height=6}
plot_residual_diag(fit_sarimax_D1_2, y_train$timestamp)
```
```{r}
plot_residual_acf_longlag(fit_sarimax_D1_2)
```


**Validation**
```{r}
weather_vars <- weather_vars_subset2

xreg_val_subset <- xreg_val[, weather_vars, drop = FALSE]
fc_sarima_D1_2 <- forecast(fit_sarimax_D1_2, h = H_val, xreg = xreg_val_subset)
```


```{r}
calc_metrics(y_val_ts, fc_sarima_D1_2$mean)
```

```{r fig.width=12, fig.height=4}
plot_train_valid_forecast(
  y_train = y_train,
  y_train_ts = y_train_ts,
  y_val = y_val,
  y_val_ts = y_val_ts,
  fc_obj = fc_sarima_D1_2,
  ci_level = 2,
  lookback_days = 7*12
)
```


### Forecast output on Test Data

#### Train the model on whole training data

```{r}
Xreg_full <- df_xreg_1819 %>% select(-timestamp) %>% as.matrix()

y_full <- ts(df_1819_cut$electricity_demand, frequency = 24)

final_fit_sarimax <- auto.arima(y_full,
  xreg = Xreg_full,
  order = c(4,0,3),
  seasonal = list(order = c(2,1,1), period = 24),
  method = "CSS-ML",
  lambda = 0
)
```

#### Forecast on test
```{r}
# forecast horizon
h_fc <- 24*7*4

Xreg_test <- df_xreg_20 %>%
  slice(seq_len(h_fc)) %>%
  select(-timestamp) %>%
  as.data.frame()

Xreg_test <- Xreg_test[, colnames(Xreg_full), drop = FALSE] %>% as.matrix()

fc_test <- forecast(final_fit_sarimax, h = h_fc, xreg = Xreg_test)

y_pred <- as.numeric(fc_test$mean)
forecast_df_test <- data.frame(
  timestamp = df_20_cut$timestamp[seq_len(h_fc)],
  Forecast  = y_pred,
  Lower     = fc_test$lower[,2],  # 95%
  Upper     = fc_test$upper[,2]
)
```


#### Save forecast results of test data

```{r}
modeling_method <- "SARIMAX"
forecast_path  <- paste0("data/test_forecast/", modeling_method, "_20Jan.csv")

write.csv(
  forecast_df_test,
  forecast_path,
  row.names = FALSE
)
```



