---
title: "Cook County Electricity Demand Time Series Modeling - SARIMAX"
output: html_document
---

```{r,warning=F,message=F}
library(forecast)
library(readr)
library(tseries)
library(ggplot2)
library(dplyr)
library(lubridate)
library(future.apply)
Sys.setlocale("LC_ALL", "en_US.UTF-8")
```

```{r}
# load data
df_1819 <- readRDS("data/train_test_split/df_1819.rds")
df_1819 <- df_1819[-1, ]
df_20   <- readRDS("data/train_test_split/df_20.rds")

cutoff <- ymd_hms("2020-01-01 00:00:00", tz="UTC")
```



### Covariate: Weather 

```{r}
weather_path <- "data/raw/ORD_weather.csv"  

df_weather <- read.csv(weather_path, stringsAsFactors = FALSE) %>%
  filter(valid != "valid") %>%
  mutate(
    timestamp = ymd_hm(valid, tz = "UTC"),
    tmpf = parse_number(tmpf),
    dwpf = parse_number(dwpf),
    sped = parse_number(sped),
    p01m = parse_number(p01m),
    ts_hour_utc = round_date(timestamp, "hour")
  )
head(df_weather,2)
```
- station: air monitoring station.
- valid: utc timestamp as string
- tmpf: air temperature [F]
- dwpf: dew point [F], the temperature to which air must be cooled to become saturated with water vapor (
 relative humidity)
- sped: wind speed [mph]
- p01m: 1 hour precipitation [mm]


**tidy up the weaather df to be aligned with elec demand df**
```{r}
# round to hour
df_weather_hr <- df_weather %>%
  group_by(ts_hour_utc) %>%
  summarise(
    across(c(tmpf, dwpf, sped, p01m),
           ~ mean(.x, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  arrange(ts_hour_utc)

# A complete hourly UTC time grid
full_hours <- tibble(
  ts_hour_utc = seq(min(df_weather_hr$ts_hour_utc),
                    max(df_weather_hr$ts_hour_utc),
                    by = "hour")
)

df_weather_full <- full_hours %>%
  left_join(df_weather_hr, by = "ts_hour_utc")

# Missing weather are linearly interpolated.
df_weather_full <- df_weather_full %>%
  arrange(ts_hour_utc) %>%
  mutate(t = as.numeric(ts_hour_utc)) %>%
  mutate(
    across(
      c(tmpf, dwpf, sped, p01m),
      ~ approx(t[!is.na(.x)],
               .x[!is.na(.x)],
               t,
               rule = 2)$y
    )
  ) %>%
  select(-t)
```

**Feature engineering on weather data**
```{r}
tmp_threshold <-50

df_weather_full <- df_weather_full %>%
  mutate(
    # time
    month = month(ts_hour_utc),
    
    # Heating Degree Hour & Cooling Degree Hour
    tmpf2 = (tmpf - tmp_threshold)^2,
    HDH   = pmax(tmp_threshold - tmpf, 0),
    HDH2  = HDH^2,
    CDH   = pmax(tmpf - tmp_threshold, 0),
    CDH2  = CDH^2,
    
    # season indicator
    is_winter = ifelse(month %in% c(12,1,2,3), 1, 0),
    is_summer = ifelse(month %in% c(6,7,8,9), 1, 0),
    
    # interaction
    winter_HDH  = HDH * is_winter,
    winter_wind = sped * is_winter,
    summer_CDH  = CDH * is_summer,
    summer_dew  = dwpf * is_summer
  )
```


### Fourier
```{r fig.width=12, fig.height=6}
y <- ts(df_1819$electricity_demand, frequency = 24)
sp <- spec.pgram(y, taper=0.1, detrend=TRUE, log="no", plot=TRUE)

abline(v = c(1, 1/7, 1/365), lty = 3, col = "red")  # daily, weekly, annual (cycles/day)
```
The x-axis frequency is cycles per day here. We can see splikes at f = 1 = Daily seasonality, f = 1/7 = Weekly seasonality, f = 1/365 ≈ 0.00274 = Annual seasonality, f ≈ 2 = 12-hour harmonic.



```{r fig.width=12, fig.height=4}
sp <- spec.pgram(y, taper=0.1, detrend=TRUE, plot=FALSE)
plot(sp$freq/24, sp$spec, type="l", xlim=c(0, 1/10),
     xlab="Frequency (cycles per hour)", ylab="Spectral density",
     main="Periodogram (zoomed: low frequencies)")
abline(v=c(1/24,1/168,1/8760), lty=3, col="red")
```
For the x-axis, frequency is divided by 24 to show seasonsality in an hourly way. spikes occur ot 1/24 and 1/168, indicating daily and weekly seasonality, while the spike at 1/12 is daily 2nd harmonic. There is also a strong low-frequency structure. might be trend and annual.

```{r}
sp <- spec.pgram(y, taper=0.1, detrend=TRUE, plot=FALSE)

# convert to cycles/hour
freq_hour <- sp$freq / 24
spec_val  <- sp$spec

# Top 10 spike
idx <- order(spec_val, decreasing = TRUE)[1:15]

data.frame(
  freq_cycles_per_hour = freq_hour[idx],
  period_hours = 1 / freq_hour[idx],
  spectral_density = spec_val[idx]
)
```

Spike at 84h is weekly harmonic. "Annual" might be mixed in the low frequencies.



### Final Xreg

```{r}
df_xreg <- df_weather_full %>%
    select(ts_hour_utc, tmpf2, HDH, CDH,winter_HDH,winter_wind, summer_CDH,p01m) %>%
    rename(timestamp = ts_hour_utc)

df_xreg_1819 <- df_xreg %>% filter(timestamp < cutoff)
df_xreg_20   <- df_xreg %>% filter(timestamp >= cutoff)
```





## Train - Validation - Test

### split train validation on 1819 data


```{r}
H_val <- 24 * 7 * 2  # 2 weeks

# 1) assure time sorting
df_1819      <- df_1819 %>% arrange(timestamp)
df_xreg_1819 <- df_xreg_1819 %>% arrange(timestamp)

# 2) Alignment check
stopifnot(nrow(df_1819) == nrow(df_xreg_1819))
stopifnot(all(df_1819$timestamp == df_xreg_1819$timestamp))

# 3) train/val index
n_1819 <- nrow(df_1819)
idx_train <- 1:(n_1819 - H_val)
idx_val   <- (n_1819 - H_val + 1):n_1819

y_train <- df_1819[idx_train, ]
y_val   <- df_1819[idx_val, ]

df_xreg_train <- df_xreg_1819[idx_train, ]
df_xreg_val   <- df_xreg_1819[idx_val, ]
```


**SARIMAX input dtype：ts + xreg matrix**
```{r}
y_col <- "electricity_demand"  

# ts: hourly, frequency=24 
y_train_msts <- msts(y_train[[y_col]], seasonal.periods = c(24, 168))
y_val_msts   <- msts(y_val[[y_col]]  , seasonal.periods = c(24, 168))
y_train_ts   <- ts(y_train[[y_col]], frequency = 24)
y_val_ts     <- ts(y_val[[y_col]]  , frequency = 24)

# xreg: drop timestamp, as.matrix
xreg_train <- df_xreg_train %>% select(-timestamp) %>% as.matrix()
xreg_val   <- df_xreg_val   %>% select(-timestamp) %>% as.matrix()

stopifnot(length(y_train_ts) == nrow(xreg_train))
stopifnot(length(y_val_ts)   == nrow(xreg_val))
stopifnot(ncol(xreg_train)   == ncol(xreg_val))
stopifnot(all(colnames(xreg_train) == colnames(xreg_val)))
```




### Assumptions made for the modeling

(S)ARIMAX models require future values of exogenous covariates when generating forecasts. In this application, weather features such as temperature and precipitation are used as an external regressor. Since reliable weather forecasts are available and generally highly accurate over short horizons, we assume that future temperature forecasts are equal to the realized temperature values. Therefore, weather features are treated as known when producing demand forecasts.

We assume the covariate can solve the heteroscedasticity issue.

The weather data is actually the records from ord station in Chicago, here we assume the weather condition in Chicago and Cook County are identical.

### Train - Validation


#### Create func for validation

```{r}
plot_residual_diag <- function(fit, timestamp) {
  res <- residuals(fit)
  par(mfrow = c(2,2), mar = c(4,4,2,1))
  plot(timestamp, res, type="l"); abline(h=0,lty=2)
  hist(res, breaks=50, prob=TRUE); lines(density(res, na.rm=TRUE))
  acf(res, lag.max=24, na.action=na.pass, ylim=c(-0.2,0.2))
  pacf(res, lag.max=24, na.action=na.pass)
  par(mfrow=c(1,1))
}

plot_residual_acf_longlag <- function(fit,ylim) {
  res <- residuals(fit)
  par(mfrow=c(2,1), mar=c(4,4,2,1))
  acf(res, lag.max=24*7*4, na.action=na.pass, ylim=c(-ylim,ylim))
  abline(v=c(7,14,21,28), col="red", lty=3)
  pacf(res, lag.max=24*7*4, na.action=na.pass)
  abline(v=c(7,14,21,28), col="red", lty=3)
  par(mfrow=c(1,1))
}
```


```{r}
calc_metrics <- function(actual, forecast_mean) {
  actual <- as.numeric(actual)
  forecast_mean <- as.numeric(forecast_mean)
  
  c(
    RMSE = sqrt(mean((actual - forecast_mean)^2)),
    MAE  = mean(abs(actual - forecast_mean)),
    MAPE = mean(abs((actual - forecast_mean) / actual)) * 100
  )
}

plot_train_valid_forecast <- function(
  y_train,           # data.frame, must contain timestamp
  y_train_ts,        # ts or numeric (train actual values)
  y_val,             # data.frame, must contain timestamp
  y_val_ts,          # ts or numeric (validation actual values)
  fc_obj,            # forecast object from forecast()
  ci_level = 2,      # which column of lower/upper to use (e.g., 2 for 95%)
  lookback_days = 7*12  # how many days before split to show
) {
  # --- forecast & actual vector ---
  yhat_valid <- as.numeric(fc_obj$mean)
  y_valid_vec <- as.numeric(y_val_ts)

  # --- align length (safety) ---
  h <- length(yhat_valid)

  # timestamps for validation/forecast must be length h
  ts_val <- y_val$timestamp[seq_len(h)]

  # --- 1. train actual ---
  train_df <- data.frame(
    timestamp = y_train$timestamp,
    value     = as.numeric(y_train_ts),
    type      = "Train"
  )

  # --- 2. validation actual (aligned) ---
  valid_df <- data.frame(
    timestamp = ts_val,
    value     = y_valid_vec[seq_len(h)],
    type      = "Validation"
  )

  # --- 3. forecast df (aligned) ---
  forecast_df <- data.frame(
    timestamp = ts_val,
    Forecast  = yhat_valid,
    Lower     = fc_obj$lower[seq_len(h), ci_level],
    Upper     = fc_obj$upper[seq_len(h), ci_level]
  )

  split_time <- min(ts_val)

  # --- 4. plot ---
  ggplot2::ggplot() +
    ggplot2::geom_line(
      data = train_df,
      ggplot2::aes(x = timestamp, y = value),
      color = "black", linewidth = 0.6
    ) +
    ggplot2::geom_line(
      data = valid_df,
      ggplot2::aes(x = timestamp, y = value),
      color = "black", linewidth = 0.6
    ) +
    ggplot2::geom_ribbon(
      data = forecast_df,
      ggplot2::aes(x = timestamp, ymin = Lower, ymax = Upper),
      fill = "pink", alpha = 0.3
    ) +
    ggplot2::geom_line(
      data = forecast_df,
      ggplot2::aes(x = timestamp, y = Forecast),
      color = "red", linewidth = 0.7
    ) +
    ggplot2::geom_vline(
      xintercept = split_time,
      linetype = "dashed", color = "blue"
    ) +
    ggplot2::theme_minimal() +
    ggplot2::labs(
      title = "Train + Validation Forecast",
      x = "Time",
      y = "Electricity Demand"
    ) +
    ggplot2::coord_cartesian(
      xlim = c(
        split_time - lubridate::days(lookback_days),
        max(ts_val)
      )
    )
}
```

#### Try 1

```{r}
weather_vars_subset1 <- c("winter_HDH","winter_wind", "summer_CDH","p01m")
Xreg_input <- xreg_train[, weather_vars_subset1, drop = FALSE]

K1 <- c(12, 6)
X_fourier <- fourier(y_train_msts, K=K1)

fit_arimax_fourier_1 <- auto.arima(
  y_train_msts,
  xreg = cbind(Xreg_input,X_fourier),
  seasonal = FALSE
)

summary(fit_arimax_fourier_1)
```

**Check Residuals**
```{r fig.width=12, fig.height=6}
plot_residual_diag(fit_arimax_fourier_1, y_train$timestamp)
```
```{r}
plot_residual_acf_longlag(fit_arimax_fourier_1)
```

Obiviously that, acf and pacf all have significant weekly lags, indicating that seasonality = 168 should be considered. 

This shows the limitation of SARIMAX can only have one seasonality.


**Validation**
```{r}
weather_vars <- weather_vars_subset1
xreg_val_subset <- xreg_val[, weather_vars, drop = FALSE]

X_fourier_val <- fourier(y_train_msts, K = K1, h = H_val)
X_val_all <- cbind(xreg_val_subset, X_fourier_val)

fc_arimax_fourier_1 <- forecast(fit_arimax_fourier_1, h = H_val, xreg = X_val_all)
```


```{r}
calc_metrics(y_val_ts, fc_arimax_fourier_1$mean)
```

```{r fig.width=12, fig.height=4}
plot_train_valid_forecast(
  y_train = y_train,
  y_train_ts = y_train_ts,
  y_val = y_val,
  y_val_ts = y_val_ts,
  fc_obj = fc_arimax_fourier_1,
  ci_level = 2,
  lookback_days = 7*12
)
```

#### Try 2: SARIMA(…)[24] + weekly Fourier

```{r}
# auto.arima -> (3,0,1)(3,1,2)[24] with "tmpf2"
weather_vars_subset2 <- c("winter_HDH", "summer_CDH","p01m")

K_week <- 3 
X_week_fourier_train <- fourier(y_train_msts, K = c(0, K_week))
Xreg_input <- cbind(xreg_train[, weather_vars_subset2, drop=FALSE], X_week_fourier_train)

fit_sarimax_D1_weekF_1 <- Arima(
  y_train_ts,
  xreg = Xreg_input,
  order = c(3,0,1),
  seasonal = list(order = c(2,1,1), period = 24),
  method = "CSS-ML",
)

summary(fit_sarimax_D1_weekF_1)
```

**Check Residuals**
```{r fig.width=12, fig.height=6}
plot_residual_diag(fit_sarimax_D1_weekF_1, y_train$timestamp)
```
```{r}
plot_residual_acf_longlag(fit_sarimax_D1_weekF_1,0.2)
```


**Validation**
```{r}
weather_vars <- weather_vars_subset2

X_week_fourier_val <- fourier(y_train_msts, K = c(0, K_week), h = H_val)

xreg_val_input <- cbind(
  xreg_val[, weather_vars, drop = FALSE],
  X_week_fourier_val
)

fc_sarimax_D1_weekF_1 <- forecast(fit_sarimax_D1_weekF_1, h = H_val, xreg = xreg_val_input)
```

```{r}
calc_metrics(y_val_ts, fc_sarimax_D1_weekF_1$mean)
```

```{r fig.width=12, fig.height=4}
plot_train_valid_forecast(
  y_train = y_train,
  y_train_ts = y_train_ts,
  y_val = y_val,
  y_val_ts = y_val_ts,
  fc_obj = fc_sarimax_D1_weekF_1,
  ci_level = 2,
  lookback_days = 7*12
)
```
#### Try 3: SARIMA(…)[24] + weekly Fourier K = 5

```{r}
# auto.arima -> (3,0,1)(3,1,2)[24] with "tmpf2"
weather_vars_subset2 <- c("winter_HDH", "summer_CDH","p01m")

K_week2 <- 5
X_week_fourier_train <- fourier(y_train_msts, K = c(0, K_week2))
Xreg_input <- cbind(xreg_train[, weather_vars_subset2, drop=FALSE], X_week_fourier_train)

fit_sarimax_D1_weekF_2 <- Arima(
  y_train_ts,
  xreg = Xreg_input,
  order = c(3,0,1),
  seasonal = list(order = c(2,1,1), period = 24),
  method = "CSS-ML",
)

summary(fit_sarimax_D1_weekF_2)
```

**Check Residuals**
```{r fig.width=12, fig.height=6}
plot_residual_diag(fit_sarimax_D1_weekF_2, y_train$timestamp)
```
```{r}
plot_residual_acf_longlag(fit_sarimax_D1_weekF_2,0.2)
```


**Validation**
```{r}
weather_vars <- weather_vars_subset2

X_week_fourier_val <- fourier(y_train_msts, K = c(0, K_week2), h = H_val)

xreg_val_input <- cbind(
  xreg_val[, weather_vars, drop = FALSE],
  X_week_fourier_val
)

fc_sarimax_D1_weekF_2 <- forecast(fit_sarimax_D1_weekF_2, h = H_val, xreg = xreg_val_input)
```

```{r}
calc_metrics(y_val_ts, fc_sarimax_D1_weekF_2$mean)
```

```{r fig.width=12, fig.height=4}
plot_train_valid_forecast(
  y_train = y_train,
  y_train_ts = y_train_ts,
  y_val = y_val,
  y_val_ts = y_val_ts,
  fc_obj = fc_sarimax_D1_weekF_2,
  ci_level = 2,
  lookback_days = 7*12
)
```

### Forecast output on Test Data

#### Train the model on whole training data

```{r}
y_1819_ts   <- ts(df_1819$electricity_demand, frequency = 24)
y_1819_msts <- msts(df_1819$electricity_demand, seasonal.periods = c(24,168))


Xreg_1819 <- df_xreg_1819 %>% select(winter_HDH, summer_CDH, p01m) %>% as.matrix()
X_week_fourier_1819 <- fourier(y_1819_msts, K = c(0, 5))
xreg_train <- cbind(Xreg_1819, X_week_fourier_1819)

stopifnot(nrow(xreg_train) == length(y_1819_ts))

final_fit_sarimax_fourier <- Arima(
  y_1819_ts,
  xreg = xreg_train,
  order = c(3,0,1),
  seasonal = list(order = c(2,1,1), period = 24),
  method = "CSS-ML"
)
```

#### Forecast on test
```{r}
# forecast horizon
h_fc <- 24*7*4

Xreg_test <- df_xreg_20 %>%
  slice(seq_len(h_fc)) %>%
  select(all_of(colnames(Xreg_1819))) %>%  
  as.matrix()

X_week_future <- fourier(y_1819_msts, K = c(0, 5), h = h_fc)
xreg_future <- cbind(Xreg_test, X_week_future)

stopifnot(nrow(xreg_future) == h_fc)

fc_test <- forecast(final_fit_sarimax_fourier, h = h_fc, xreg = xreg_future)

forecast_df_test <- data.frame(
  timestamp = df_xreg_20$timestamp[seq_len(h_fc)],
  Forecast  = as.numeric(fc_test$mean),
  Lower     = fc_test$lower[,2],  # 95%
  Upper     = fc_test$upper[,2]
)
```


#### Save forecast results of test data

```{r}
modeling_method <- "SARIMAX_w_fourier"
forecast_path  <- paste0("data/test_forecast/", modeling_method, "_20Jan.csv")

write.csv(
  forecast_df_test,
  forecast_path,
  row.names = FALSE
)
```



