---
title: "Cook County Electricity Demand Time Series Modeling - SARIMAX"
output: html_document
---

```{r,warning=F,message=F}
library(forecast)
library(tseries)
library(ggplot2)
library(dplyr)
library(lubridate)
library(future.apply)
Sys.setlocale("LC_ALL", "en_US.UTF-8")
```

```{r}
# load data
df_1819 <- readRDS("data/train_test_split/df_1819.rds")
df_20   <- readRDS("data/train_test_split/df_20.rds")
```

### Covariate: temperature

```{r}
temp_path <- "data/raw/ord_temp.txt"  

df_temp <- read.csv(
  temp_path,
  header = FALSE,
  col.names = c("station", "timestamp", "temperature"),
  stringsAsFactors = FALSE
)
df_temp <- df_temp %>% filter(station != "station")

# Local Chicago timestamps are converted to UTC to match the load data.
df_temp$timestamp <- as.POSIXct(
  df_temp$timestamp,
  format = "%Y-%m-%d %H:%M",
  tz = "America/Chicago"
)

df_temp <- df_temp %>% mutate( timestamp_utc = with_tz(timestamp, "UTC") )

range(df_temp$timestamp_utc)
```

```{r}
library(readr)

# Temperatures are converted to numeric and rounded to the nearest hour.
df_temp2 <- df_temp %>%
  mutate(
    temperature = parse_number(temperature),  
    ts_hour_utc = round_date(timestamp_utc, "hour")
  )

df_temp_hr <- df_temp2 %>%
  group_by(ts_hour_utc) %>%
  summarise(temp = mean(temperature, na.rm = TRUE), .groups = "drop") %>%
  arrange(ts_hour_utc)

# A complete hourly UTC time grid
full_hours <- tibble(
  ts_hour_utc = seq(
    min(df_temp_hr$ts_hour_utc),
    max(df_temp_hr$ts_hour_utc),
    by = "hour"
  )
)

df_temp_full <- full_hours %>% left_join(df_temp_hr, by = "ts_hour_utc")

df_temp_full <- df_temp_full %>%
  mutate(t_num = as.numeric(ts_hour_utc))


# Missing temperatures are linearly interpolated.
ok <- !is.na(df_temp_full$temp)

df_temp_full$temp <- approx(
  x = df_temp_full$t_num[ok],
  y = df_temp_full$temp[ok],
  xout = df_temp_full$t_num,
  method = "linear",
  rule = 2
)$y

df_temp_full <- df_temp_full %>%
  select(ts_hour_utc, temp)
```
```{r fig.width=12, fig.height=4}
ggplot(df_temp_full, aes(x = ts_hour_utc, y = temp)) +
  geom_line(color = "steelblue") +
  labs(
    title = "Hourly Temperature (ORD)",
    x = "Time (UTC)",
    y = "Temperature (°F)"
  ) +
  theme_minimal()
```



## Train - Validation - Test

### Prepare Train - test data

Use the time range of temperature to crop the time range of load (ensuring they are completely aligned)

```{r}
### split train test 
temp_start <- min(df_temp_full$ts_hour_utc)
temp_end   <- max(df_temp_full$ts_hour_utc)
# elec demand
df_1819_cut <- df_1819 %>% filter(timestamp >= temp_start)
df_20_cut <- df_20 %>% filter(timestamp <= temp_end)
# temp
cutoff <- ymd_hms("2020-01-01 00:00:00", tz = "UTC")
temp_1819 <- df_temp_full %>% filter(ts_hour_utc >= temp_start, ts_hour_utc < cutoff)
temp_20 <- df_temp_full %>% filter(ts_hour_utc >= cutoff, ts_hour_utc <= temp_end)


### split train - validation on "train df"
# create aligned train dataframe
train_xy <- df_1819_cut %>%
  select(timestamp, y = electricity_demand) %>%
  inner_join(temp_1819, by = c("timestamp" = "ts_hour_utc")) %>%
  arrange(timestamp)
# train - valid split
H_val <- 56 * 24
n <- nrow(train_xy)
train_part <- train_xy[1:(n - H_val), ]
valid_part <- train_xy[(n - H_val + 1):n, ]
# Xreg
center_temp = 45
temp_center_train <- train_part$temp - center_temp
X_train <- cbind(temp  = temp_center_train,temp2 = temp_center_train^2)
temp_center_valid <- valid_part$temp - center_temp
X_valid <- cbind(temp  = temp_center_valid,temp2 = temp_center_valid^2)

# convert to expected input dtype
y_train <- ts(train_part$y, frequency = 24)
y_valid <- ts(valid_part$y, frequency = 24)
```

```{r fig.width=12, fig.height=6}
plot_df <- train_part %>%
  mutate(
    temp_center = temp - center_temp,
    temp_sq     = temp_center^2
  ) %>%
  select(timestamp, y, temp_center, temp_sq) %>%
  tidyr::pivot_longer(
    cols = c(y, temp_center, temp_sq),
    names_to = "series",
    values_to = "value"
  )

ggplot(plot_df, aes(x = timestamp, y = value)) +
  geom_line() +
  facet_wrap(~series, ncol = 1, scales = "free_y") +
  theme_minimal() +
  labs(title = "Train Data: Demand vs Centered Temperature Features",
       x = NULL,
       y = NULL)
```


### Assumptions made for the modeling

(S)ARIMAX models require future values of exogenous covariates when generating forecasts. In this application, temperature is used as an external regressor. Since reliable weather forecasts are available and generally highly accurate over short horizons, we assume that future temperature forecasts are equal to the realized temperature values. Therefore, temperature is treated as known when producing demand forecasts.

We assume the covriate can solve the heteroscedasticity issue.

The temperature data is the record from station in Chicago, here we assume the weather condition in Chicago and Cook County are identical.

### Train - Validation

#### Try 1

```{r}
# auto.arima -> ARIMA(1,1,0)(2,1,2)[24]
fit_arima_quick <- auto.arima(
  y_train,
  xreg = X_train[, "temp2", drop = FALSE],
  seasonal = TRUE,
  stepwise = FALSE,
  approximation = TRUE,
  d = 1, D = 1,
  max.p = 4, max.q = 4,
  max.P = 3, max.Q = 3
)
summary(fit_arima_quick)
```

```{r fig.width=12, fig.height=6}
# Residual diagnostics
res <- residuals(fit_arima_quick)

par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))

# (1) Residuals over time
plot(train_part$timestamp, res, type = "l",
     main = "Residuals (Time Plot)", xlab = "Time", ylab = "Residual")
abline(h = 0, lty = 2)

# (2) Residual distribution
hist(res, breaks = 50, probability = TRUE,
     main = "Residual Distribution", xlab = "Residual")
lines(density(res, na.rm = TRUE))

# (3) ACF short (0–24 lags)
acf(res, lag.max = 24, na.action = na.pass, ylim = c(-0.2, 0.2))

# (4) PACF short (0–24 lags)
pacf(res, lag.max = 24, na.action = na.pass)

par(mfrow = c(1, 1))
```
```{r}
par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))

# (5) ACF long  (0–4 weeks)
acf(res, lag.max = 24 * 7 * 4, na.action = na.pass, ylim = c(-0.2, 0.2))
abline(v = c(7,14,21,28), col = "red", lty = 3)

# (6) PACF long (0–4 weeks)
pacf(res, lag.max = 24 * 7 * 4, na.action = na.pass)
abline(v = c(7,14,21,28), col = "red", lty = 3)

```


#### Validation 1
```{r}
fc <- forecast(fit_arima_quick, h = H_val, xreg = X_valid[, "temp2", drop = FALSE])

yhat_valid <- as.numeric(fc$mean)
y_valid_vec <- as.numeric(y_valid)

rmse <- function(a, f) sqrt(mean((a - f)^2, na.rm = TRUE))
mae  <- function(a, f) mean(abs(a - f), na.rm = TRUE)
mape <- function(a, f) mean(abs((a - f) / a), na.rm = TRUE) * 100

valid_RMSE <- rmse(y_valid_vec, yhat_valid)
valid_MAE  <- mae(y_valid_vec, yhat_valid)
valid_MAPE <- mape(y_valid_vec, yhat_valid)

valid_RMSE
valid_MAE
valid_MAPE
```



#### Try 2
```{r}
fit_sarimax <- Arima(y_train,
  xreg = X_train[, "temp2", drop = FALSE],
  order = c(4,0,3),
  seasonal = list(order = c(2,1,1), period = 24),
  method = "CSS-ML",
  lambda = 0
)
summary(fit_sarimax)
```

```{r fig.width=12, fig.height=6}
# Residual diagnostics
res2 <- residuals(fit_sarimax)

par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))

# (1) Residuals over time
plot(train_part$timestamp, res2, type = "l",
     main = "Residuals (Time Plot)", xlab = "Time", ylab = "Residual")
abline(h = 0, lty = 2)

# (2) Residual distribution
hist(res2, breaks = 50, probability = TRUE,
     main = "Residual Distribution", xlab = "Residual")
lines(density(res2, na.rm = TRUE))

# (3) ACF short (0–24 lags)
acf(res2, lag.max = 24, na.action = na.pass, ylim = c(-0.2, 0.2))

# (4) PACF short (0–24 lags)
pacf(res2, lag.max = 24, na.action = na.pass)

par(mfrow = c(1, 1))
```
```{r}
par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))

# (5) ACF long  (0–4 weeks)
acf(res2, lag.max = 24 * 7 * 4, na.action = na.pass, ylim = c(-0.2, 0.2))
abline(v = c(7,14,21,28), col = "red", lty = 3)

# (6) PACF long (0–4 weeks)
pacf(res2, lag.max = 24 * 7 * 4, na.action = na.pass)
abline(v = c(7,14,21,28), col = "red", lty = 3)

```


#### Validation 2

```{r}
fc2 <- forecast(fit_sarimax, h = H_val, xreg = X_valid[, "temp2", drop = FALSE])

yhat_valid2 <- as.numeric(fc2$mean)
y_valid_vec <- as.numeric(y_valid)

valid_RMSE <- rmse(y_valid_vec, yhat_valid2)
valid_MAE  <- mae(y_valid_vec, yhat_valid2)
valid_MAPE <- mape(y_valid_vec, yhat_valid2)

valid_RMSE
valid_MAE
valid_MAPE
```
```{r fig.width=12, fig.height=6}
yhat_valid_plot <- yhat_valid2
fc_plot <-fc2

# --- 1. train actual ---
train_df <- data.frame(
  timestamp = train_part$timestamp,
  value     = as.numeric(y_train),
  type      = "Train"
)

# --- 2. validation actual ---
valid_df <- data.frame(
  timestamp = valid_part$timestamp,
  value     = y_valid_vec,
  type      = "Validation"
)

# --- 3. forecast ---
forecast_df <- data.frame(
  timestamp = valid_part$timestamp,
  Forecast  = yhat_valid_plot,
  Lower     = fc_plot$lower[,2],
  Upper     = fc_plot$upper[,2]
)

# --- 4. plot ---
ggplot() +
  # Train actual
  geom_line(data = train_df,
            aes(x = timestamp, y = value),
            color = "black", linewidth = 0.6) +
  
  # Validation actual
  geom_line(data = valid_df,
            aes(x = timestamp, y = value),
            color = "black", linewidth = 0.6) +
  
  # Forecast CI
  geom_ribbon(data = forecast_df,
              aes(x = timestamp, ymin = Lower, ymax = Upper),
              fill = "pink", alpha = 0.3) +
  
  # Forecast line
  geom_line(data = forecast_df,
            aes(x = timestamp, y = Forecast),
            color = "red", linewidth = 0.7) +
  
  # vertical split line
  geom_vline(xintercept = min(valid_part$timestamp),
             linetype = "dashed", color = "blue") +
  
  theme_minimal() +
  labs(title = "Train + Validation Forecast",
       x = "Time",
       y = "Electricity Demand")

coord_cartesian(xlim = c(min(valid_part$timestamp) - days(24*7*4*6),
                         max(valid_part$timestamp)))
```


### Forecast output on Test Data

#### Train the model on whole training data

```{r}
# creat xreg
temp_center_full <- (temp_1819$temp - 40)^2
X_full <- cbind(temp2 = temp_center_full)

y_full <- ts(df_1819_cut$electricity_demand, frequency = 24)

final_fit_sarimax <- auto.arima(y_full,
  xreg = X_full,
  order = c(4,0,3),
  seasonal = list(order = c(2,1,1), period = 24),
  method = "CSS-ML",
  lambda = 0
)
```

#### Forecast on test
```{r}
temp_center_test <- (temp_20$temp - 40)^2
X_test <- cbind(temp2 = temp_center_test)

h_fc <- nrow(df_20_cut)

fc_test <- forecast(
  final_fit_sarimax,
  h = h_fc,
  xreg = X_test
)

y_pred <- as.numeric(fc_test$mean)
forecast_df_test <- data.frame(
  timestamp = df_20_cut$timestamp,
  Forecast  = y_pred,
  Lower     = fc_test$lower[,2],  # 95%
  Upper     = fc_test$upper[,2]
)
```


#### Save forecast results of test data

```{r}
modeling_method <- "SARIMAX"
forecast_path  <- paste0("data/test_forecast/", modeling_method, "_20Jan.csv")

write.csv(
  forecast_df_test,
  forecast_path,
  row.names = FALSE
)
```



