---
title: "TS_Final_Project_Preprocessing"
author: "Jenny Dong"
date: "`r Sys.Date()`"
output: html_document
---

# Load libraries

```{r,warning=F,message=F}
library(forecast)
library(tseries)
library(ggplot2)
library(dplyr)
Sys.setlocale("LC_ALL", "en_US.UTF-8")
```

# Data Understanding

### Loading data

Due to software and memory limitations, the dataset was lightly cleaned and filtered to include only 2018/1 \~ 2020/1 in Cook County, Chicago, IL (p17031) in Python prior to being exported for time series analysis in R.

```{r}
file_path <- "cook_county_2018_to_2020Jan.csv"
df <- read.csv(file_path)
colnames(df) <- c("timestamp", "electricity_demand")
```

electricity demand from 2018/1 to 2020/1 in Cook County:

```{r}
glimpse(df)
```

### Preprocessing

#### timestamp conversion:

```{r}
library(lubridate)
# The datetime64[ns, tz] information stored in the original data HDF5 shows as timezone-aware UTC DatetimeIndex
df$timestamp_utc <- ymd_hms(df$timestamp, tz = "UTC")
df$timestamp_cst <- with_tz(df$timestamp_utc, "America/Chicago")
head(df,5)
```

Since the CST time zone involves the conversion of daylight saving time, during the analysis we maintain the use of a more standardized UTC, and the explanatory analysis can be converted back to CST.

```{r}
# Create a new clean dataframe
df_cleaned <- data.frame(
  timestamp = df$timestamp_utc,
  electricity_demand = df$electricity_demand
)
head(df_cleaned,5)
```

#### quick view:

```{r}
library(dplyr)
glimpse(df_cleaned)
```

```{r}
library(skimr)
skim(df_cleaned)
```

#### split data

```{r}
cutoff <- ymd_hms("2020-01-01 00:00:00", tz="UTC")

df_1819 <- df_cleaned %>% filter(timestamp < cutoff)
df_20   <- df_cleaned %>% filter(timestamp >= cutoff)

cat(sprintf(
  "Train: %s → %s | Test: %s → %s\n",
  format(min(df_1819$timestamp), "%Y-%m-%d %H:%M:%S"),
  format(max(df_1819$timestamp), "%Y-%m-%d %H:%M:%S"),
  format(min(df_20$timestamp),   "%Y-%m-%d %H:%M:%S"),
  format(max(df_20$timestamp),   "%Y-%m-%d %H:%M:%S")
))
```

```{r}
# Build multi-seasonal time series
ts_1819 <- msts(df_1819$electricity_demand, seasonal.periods = c(24, 24*7, 24*365.25))
ts_20   <- msts(df_20$electricity_demand,   seasonal.periods = c(24, 24*7, 24*365.25))
```

### EDA

#### plot ts

```{r fig.width=12, fig.height=5}
# Plot Hourly Electricity Demand of 2018 ~ 2019
ggplot(df_1819, aes(x = timestamp, y = electricity_demand)) +
  geom_line(color = "deepskyblue") +
  labs(title = "Hourly Electricity Demand (2018/1 ~ 2019/12)",
       x = "Time",
       y = "Electricity Demand") +
  theme_minimal()
```

```{r fig.width=12, fig.height=4}
ggplot(df_1819[df_1819$timestamp < "2018-02-01", ],
       aes(x = timestamp, y = electricity_demand)) +
  geom_line(color = "steelblue") +
  labs(title = "Hourly Electricity Demand (Jan 2018)",
       x = "Time",
       y = "Electricity Demand") +
  theme_minimal()
```

-   Clear strong seasonality (repeating cycles)
-   Noticeable variance changes (spikes in certain periods)
-   No obvious long-term upward trend
-   Large recurring peaks, daily + seasonal demand patterns

#### Transforming Need Check

```{r}
BoxCox.lambda(df_1819)
```

Since the estimated Box-Cox lambda is approximately 1 and the time series plot does not exhibit a clear pattern of increasing or decreasing variance over time, a Box-Cox transformation is not necessary.

However, BoxCox.lambda() does not "test for heteroscedasticity". We need to do Level–Scale plot to see whether a log transformation is needed.

```{r}
x <- df_1819$electricity_demand

w <- 24*7  # weekly window
mu <- stats::filter(x, rep(1/w, w), sides = 2)                 # rolling mean
sd <- sqrt(stats::filter((x - mu)^2, rep(1/w, w), sides = 2))  # rolling sd

tmp <- tibble(mu = as.numeric(mu), sd = as.numeric(sd)) %>%
  filter(is.finite(mu), is.finite(sd)) %>%
  mutate(bin = ntile(mu, 30)) %>%                 # 30 bins
  group_by(bin) %>%
  summarise(mu_bin = median(mu), sd_bin = median(sd), .groups="drop")

# 1) original scale: plot sd against mean
ggplot(tmp, aes(mu_bin, sd_bin)) +
  geom_point() + geom_smooth(se = FALSE) +
  labs(x="Local mean (level)", y="Local sd (scale)",
       title="Level–Scale (mean vs sd)")

# 2) log–log：slope≈1 => sd ∝ mean
ggplot(tmp, aes(log(mu_bin), log(sd_bin))) +
  geom_point() + geom_smooth(se = FALSE) +
  labs(x="log(Local mean)", y="log(Local sd)",
       title="Log–Log Level–Scale")

fit_alpha <- lm(log(sd_bin) ~ log(mu_bin), data = tmp)
summary(fit_alpha)$coef
```

The log-log graph is flat in low range and rises sharply in the high range in a Nonlinear pattern. Also, α = 2.58 \>\> 1. (α ≈ 1 -\> Proportional heteroscedasticity)

So, This is not a simple proportional heteroscedasticity, it's a structural heteroskedasticity.

**In this case, variance shoule be carefully checked in the following model fitted residuals. Building models for different seasons is also suggested.**

#### ACF & PACF Plots

```{r fig.width=12, fig.height=4}
acf(ts_1819, main = "ACF of Training")
```

-   Lag 1 is very close to 1.
-   ACF decays very slow in a seasonal wavy structure, with many lags far exceeding the significance limit,
-   strong non-stationarity.

**The hourly electricity demand data is probably a superimposed structure of** $Y_t = Trend + Season_{24} + Season_{168} + Season_{8760} + ε_t$


```{r fig.width=12, fig.height=4}
ndiffs(ts_1819)      # regular difference

ts_d1_1819 <- diff(ts_1819)
acf(ts_d1_1819)
```
A differecing  = 1 is a must here to remove the stochastic trend here. We have no knowledge and confidence to determine a deterministic trend in this case.

```{r fig.width=12, fig.height=5}
ts_d1_D24_1819  <- diff(ts_d1_1819, lag = 24) 
acf(ts_d1_D24_1819, lag.max=168)
pacf(ts_d1_D24_1819, lag.max=168)
```
```{r fig.width=12, fig.height=5}
ts_d1_D168_1819  <- diff(ts_d1_1819, lag = 24*7) 
acf(ts_d1_D168_1819, lag.max=168)
pacf(ts_d1_D168_1819, lag.max=168)
```

Just a quick look here. Seasonal differencing will be further determined in the SARIMA modeling part.



The HEGY test result in p-values of 0 across all regular and seasonal harmonic frequencies, conclusively rejecting the null hypothesis of unit roots. The hourly electricity demand time series is strictly stationary around its deterministic constant and trend. Consequently, the data requires no further regular or seasonal differencing before proceeding to the SARIMAX and ETS modeling phases.

#### STL Decomposition

```{r fig.width=12, fig.height=5}
mstl_decomp_1819 <- mstl(ts_1819)
plot(mstl_decomp_1819)
```

##### Check Residuals from STL Decomposition

```{r fig.width=12, fig.height=8}
rem <- mstl_decomp[, "Remainder"]
par(mfrow=c(3,1), mar=c(3,4,2,1))
plot(rem, main="MSTL Remainder", ylab="Residual")
acf(rem, main="ACF of Remainder", lag.max = 24*14)   # 看两周
pacf(rem, main="PACF of Remainder", lag.max = 24*14)
par(mfrow=c(1,1))
```

Check residual Normality:

```{r  fig.height=4}
hist(rem, main = "Histogram of MSTL Residuals", breaks = 40)
```

Ljung-Box Test: H0: Residuals are independent vs. Ha: Residuals are autocorrelated (bad)

```{r}
Box.test(rem, lag = 24, type = "Ljung-Box")
```

# Modeling

### Simple Methods (Baselines)

```{r}
# train
ts_1819 <- ts(df_1819$electricity_demand, frequency = 24)

# test
ts_2020 <- ts(
  df_2020$electricity_demand,
  frequency = 24,
  start = time(ts_1819)[length(ts_1819)] + 1/frequency(ts_1819)
)

# seasonal naive
fit_snaive <- snaive(ts_1819, h = length(ts_2020))

# plot
autoplot(fit_snaive) +
  autolayer(ts_2020, series = "Actual Demand (Jan 2020)", color = "red") +
  coord_cartesian(xlim = c(max(time(ts_1819)) - 10, max(time(ts_2020)))) +
  labs(title = "Seasonal Naive Forecast vs Actual",
       x = "Time (Days)",
       y = "Electricity Demand") +
  theme_minimal()

# accuracy
acc_snaive <- accuracy(fit_snaive, ts_2020)
print(acc_snaive)

checkresiduals(fit_snaive)
```

The plot compares this repeated daily profile against the January 2020 actual demand, so any persistent gap indicates level changes or calendar/weather-driven shifts not present at the end of 2019. Quantitatively, the test MAPE is about 5.25%, which is a reasonable benchmark, but the very high residual autocorrelation suggests substantial remaining structure beyond simple daily seasonality.

```{r}
# Fit Mean Forecast model
fit_mean <- meanf(ts_1819, h = length(ts_2020))

# Plot the Mean forecast
autoplot(fit_mean, PI = FALSE, series = "Mean Forecast") +
  autolayer(ts_2020, series = "Actual Demand (Jan 2020)", color = "red") +
  coord_cartesian(xlim = c(max(time(ts_1819)) - 10, max(time(ts_2020)))) +
  labs(title = "Mean Forecast vs Actual",
       x = "Time (Days)",
       y = "Electricity Demand") +
  theme_minimal() +
  scale_color_manual(values = c("Actual Demand (Jan 2020)" = "red", "Mean Forecast" = "blue"))

# Evaluate accuracy bypassing ts index mismatch
acc_mean <- accuracy(fit_mean, df_2020$electricity_demand)
print(acc_mean)

```

```{r}
# Fit STLF model
fit_stlf <- stlf(ts_1819, method = "naive", h = length(ts_2020))

# Plot the STLF forecast
autoplot(fit_stlf, PI = FALSE, series = "STLF Forecast") +
  autolayer(ts_2020, series = "Actual Demand (Jan 2020)", color = "red") +
  coord_cartesian(xlim = c(max(time(ts_1819)) - 10, max(time(ts_2020)))) +
  labs(title = "STLF Forecast vs Actual",
       x = "Time (Days)",
       y = "Electricity Demand") +
  theme_minimal() +
  scale_color_manual(values = c("Actual Demand (Jan 2020)" = "red", "STLF Forecast" = "blue"))

# Evaluate accuracy
acc_stlf <- accuracy(fit_stlf, df_2020$electricity_demand)
print(acc_stlf)

```

```{r}
#Time Series Linear Model (TSLM) & Diagnostics
library(lmtest)

# Fit TSLM with trend and daily seasonality
fit_tslm <- tslm(ts_1819 ~ trend + season)
summary(fit_tslm)

autoplot(window(ts_1819, start = time(ts_1819)[length(ts_1819) - 336])) +
  autolayer(window(fitted(fit_tslm), start = time(ts_1819)[length(ts_1819) - 336]), 
            series = "Fitted (TSLM)", size = 0.8) +
  labs(title = "TSLM: Actual vs. Fitted (Last 14 Days of Training)",
       x = "Time",
       y = "Electricity Demand") +
  theme_minimal() +
  scale_color_manual(values = c("Data" = "black", "Fitted (TSLM)" = "red"))

# Residual Diagnostics
res_tslm <- residuals(fit_tslm)

# Plot Residuals Time Plot
autoplot(res_tslm) +
  labs(title = "Residuals from TSLM", y = "Residuals") +
  theme_minimal()

# Plot ACF of Residuals
ggAcf(res_tslm, lag.max = 48) +
  labs(title = "ACF of TSLM Residuals")

# Plot Histogram of Residuals
hist(res_tslm, breaks = 50, col = "lightblue", 
     main = "Histogram of TSLM Residuals", xlab = "Residuals")

# Ljung-Box Test
Box.test(res_tslm, lag = 24, type = "Ljung-Box")

# Breusch-Godfrey Test
bgtest(fit_tslm)

# Forecast and Plot
fcast_tslm <- forecast(fit_tslm, h = length(df_2020$electricity_demand))

autoplot(fcast_tslm, PI = FALSE, series = "TSLM Forecast") +
  autolayer(ts_2020, series = "Actual Demand (Jan 2020)") +
  coord_cartesian(xlim = c(max(time(ts_1819)) - 10, max(time(ts_2020)))) +
  labs(title = "TSLM Forecast vs Actual",
       x = "Time (Days)",
       y = "Electricity Demand") +
  theme_minimal() +
  scale_color_manual(values = c("Actual Demand (Jan 2020)" = "red", "TSLM Forecast" = "blue"))

acc_tslm <- accuracy(fcast_tslm, df_2020$electricity_demand)
print(acc_tslm)
```

For visualization purposes, only the last 14 days of the testing period are shown. This window includes multiple daily and weekly seasonal cycles, making it sufficient to evaluate forecast alignment while maintaining visual clarity.
